```{r, echo=FALSE, results='hide', message=FALSE, warning=FALSE}
load("C:/Users/Joshua/Desktop/Australian National University/2nd year/Semester 2/STAT3011- Graphical Data Analysis/classdata (1).RData")
insure <- as.data.frame(insure)
attach(insure)
library(psych)
library(ggplot2)
library(RColorBrewer)
library(usethis)
library(Rcpp)
library(devtools)
library(choroplethrZip)
library(pander)
library(gridExtra)
library(knitr)
```

##Insurance availability in Chicago

>The U.S. Commission on Civil Rights collected data in an attempt to
examine charges that insurance companies were "redlining" certain
neighbourhoods.  i.e.  cancelling and/or refusing to renew policies.
The data provided include the number of cancellations, nonrenewals, new
policies and renewals of home and fire policies for each neighbourhood
by zip code for the months December 1977 - February 1978.  This
information is combined into a single variable denoted Voluntary market
activity which is the number of new policies and renewals minus the
number of cancellations and nonrenewals expressed per 100 housing
units.  In addition, information on the number of FAIR plan policies
was obtained.  These policies are obtained after applicants have been
rejected for other policies so this information also reflects the
availability of policies.  This information is provided as the
involuntary market activity, the number of FAIR plan policies and
renewals per 100 housing units.  In addition, the Chicago Police
provided theft data and the Fire Department provided fire data from
1975 for each neighbourhood.  These data are the number of incidents
per 1000 housing units in 1975.  (The insurance companies claim to use
a three year lag on crime data when they set their premiums.) Finally,
the Census Bureau provide data on the racial composition (in per cent
minority), income and the age of housing units.  The income is the
median family income and the age is coded as the percentage of units
built in or before 1939.

>The objectives of the study are to explore the extent to which racial
composition and age of housing affect underwriting practices after
controlling for factors like fire and theft.

```{r, echo=FALSE,  message=FALSE, warning=FALSE, fig.width=9, fig.height=9, fig.align='center', fig.cap="Scatterplot matrix of Chicago insurance data, showcasing the bivariate relationships between voluntary and involuntary insurance market activity and factors such as theft and fire incidents, age of housing units, median income and percentage of minority racial composition."}
pairs.panels(insure, method = 'pearson', density = TRUE , smooth = TRUE, ellipses = FALSE, cor = TRUE, scale = TRUE, cex.cor = 2, hist.col = '#00AFBB', rug = FALSE)
```

```{r, echo=FALSE,  message=FALSE, warning=FALSE, results=FALSE}
a0 <- as.character(insure$Zip)
a1 <- as.numeric(c(insure$Volun, insure$Invol))
a2 <- as.factor(c(rep('2',length(insure$Volun)), rep('1', length(insure$Invol))))
policy.data <- as.data.frame(cbind(a0,a2,a1))
colnames(policy.data) <- c('Zip code', 'Policy type', 'Active policies per 100 housing units')
policy.data$`Policy type` <- as.factor(policy.data$`Policy type`)
levels(policy.data$`Policy type`)[levels(policy.data$`Policy type`)==2] <- 'Voluntary'
levels(policy.data$`Policy type`)[levels(policy.data$`Policy type`)==1] <- 'Involuntary'
policy.data$`Zip code` <- as.factor(policy.data$`Zip code`)
policy.data$`Active policies per 100 housing units` <- as.numeric(policy.data$`Active policies per 100 housing units`)
View(policy.data)
```


```{r, echo=FALSE,  message=FALSE, warning=FALSE}
ggplot(policy.data, aes(x = reorder(policy.data$`Zip code`, policy.data$`Active policies per 100 housing units`), y = policy.data$`Active policies per 100 housing units`, fill = policy.data$`Policy type`)) + geom_bar(alpha = 0.5, lwd = 0.2, position=position_dodge(), stat="identity") + labs(x = 'Zip Code') + ggtitle('Active Policies Per 100 Households') + theme_classic() + theme(legend.position = 'bottom', legend.title = element_blank(), plot.title = element_text(hjust = 0.5), legend.spacing.x = unit(0.5, 'cm'), axis.title.y = element_blank(), axis.text.y = element_blank(), axis.line.y = element_blank(), axis.ticks.y = element_blank(), axis.ticks.x = element_line(size = 0.3), axis.line.x = element_line(size = 0.3)) +  theme(axis.text.x = element_text(angle = 90, hjust = 1)) + theme(axis.text.y = element_text(angle = 90, hjust = 1))

```

>Figure 2: Comparison between voluntary and involuntary insurance market activity. Involuntary market
activity is the number of new policies and renewals minus the
number of cancellations and nonrenewals expressed per 100 housing
units. Involuntary market activity is the number of FAIR plan policies and
renewals per 100 housing units.

```{r, echo=FALSE,  message=FALSE, warning=FALSE, fig.width=10, fig.height=16}

col1.0 <- insure$Zip
col2.0 <- insure$Theft
theft.data <- as.data.frame(cbind(col1.0,col2.0))
colnames(theft.data) <- c('region', 'value')
theft.data$region <- as.character(theft.data$region)
theft.data$value <- as.numeric(theft.data$value)
theft.data[18,1] <- '60827'
theft.data[45,1] <- '60707'

choro.theft <- ZipChoropleth$new(theft.data)
choro.theft$ggplot_scale <- scale_fill_brewer(name='Thefts per 1000 houses', palette = 'BuPu', drop=FALSE)
choro.theft$set_zoom_zip(state_zoom =NULL, county_zoom=NULL, msa_zoom=NULL, zip_zoom = c('60607', '60608', '60609', '60610', '60611', '60612', '60613', '60614',
                            '60615', '60616', '60617', '60618', '60619', '60620', '60621', '60622',
                            '60623', '60624', '60625', '60626', '60628', '60629', '60630', '60827',
                            '60631', '60632', '60633', '60634', '60636', '60637', '60638', '60707',
                            '60639', '60640', '60641', '60643', '60644', '60645', '60646', '60647',
                            '60649', '60651', '60652', '60653', '60655', '60656', '60657'))

col1 <- insure$Zip
col2 <- insure$Fire
fire.data <- as.data.frame(cbind(col1,col2))
colnames(fire.data) <- c('region', 'value')
fire.data$region <- as.character(fire.data$region)
fire.data$value <- as.numeric(fire.data$value)
fire.data[18,1] <- '60827'
fire.data[45,1] <- '60707'

choro.fire <- ZipChoropleth$new(fire.data)
choro.fire$ggplot_scale <- scale_fill_brewer(name='Fires per 1000 houses', palette = 'YlOrRd', drop=FALSE)
choro.fire$set_zoom_zip(state_zoom =NULL, county_zoom=NULL, msa_zoom=NULL, zip_zoom = c('60607', '60608', '60609', '60610', '60611', '60612', '60613', '60614',
                            '60615', '60616', '60617', '60618', '60619', '60620', '60621', '60622',
                            '60623', '60624', '60625', '60626', '60628', '60629', '60630', '60827',
                            '60631', '60632', '60633', '60634', '60636', '60637', '60638', '60707',
                            '60639', '60640', '60641', '60643', '60644', '60645', '60646', '60647',
                            '60649', '60651', '60652', '60653', '60655', '60656', '60657'))

col1.2 <- insure$Zip
col2.2 <- insure$Income
inc.data <- as.data.frame(cbind(col1.2,col2.2))
colnames(inc.data) <- c('region', 'value')
inc.data$region <- as.character(inc.data$region)
inc.data$value <- as.numeric(inc.data$value)
inc.data[18,1] <- '60827'
inc.data[45,1] <- '60707'

choro.inc <- ZipChoropleth$new(inc.data)
choro.inc$ggplot_scale <- scale_fill_brewer(name='Median household income', palette = 'Blues', drop=FALSE, direction = -1)
choro.inc$set_zoom_zip(state_zoom =NULL, county_zoom=NULL, msa_zoom=NULL, zip_zoom = c('60607', '60608', '60609', '60610', '60611', '60612', '60613', '60614',
                            '60615', '60616', '60617', '60618', '60619', '60620', '60621', '60622',
                            '60623', '60624', '60625', '60626', '60628', '60629', '60630', '60827',
                            '60631', '60632', '60633', '60634', '60636', '60637', '60638', '60707',
                            '60639', '60640', '60641', '60643', '60644', '60645', '60646', '60647',
                            '60649', '60651', '60652', '60653', '60655', '60656', '60657'))
col1.3 <- insure$Zip
col2.3 <- insure$Race
race.data <- as.data.frame(cbind(col1.3,col2.3))
colnames(race.data) <- c('region', 'value')
race.data$region <- as.character(race.data$region)
race.data$value <- as.numeric(race.data$value)
race.data[18,1] <- '60827'
race.data[45,1] <- '60707'

choro.race <- ZipChoropleth$new(race.data)
choro.race$ggplot_scale <- scale_fill_brewer(name='Per cent minority', palette = 'PuRd', drop=FALSE)
choro.race$set_zoom_zip(state_zoom =NULL, county_zoom=NULL, msa_zoom=NULL, zip_zoom = c('60607', '60608', '60609', '60610', '60611', '60612', '60613', '60614',
                            '60615', '60616', '60617', '60618', '60619', '60620', '60621', '60622',
                            '60623', '60624', '60625', '60626', '60628', '60629', '60630', '60827',
                            '60631', '60632', '60633', '60634', '60636', '60637', '60638', '60707',
                            '60639', '60640', '60641', '60643', '60644', '60645', '60646', '60647',
                            '60649', '60651', '60652', '60653', '60655', '60656', '60657'))



col1.1 <- insure$Zip
col2.1 <- insure$Age
age.data <- as.data.frame(cbind(col1.1,col2.1))
colnames(age.data) <- c('region', 'value')
age.data$region <- as.character(age.data$region)
age.data$value <- as.numeric(age.data$value)
age.data[18,1] <- '60827'
age.data[45,1] <- '60707'

choro.age <- ZipChoropleth$new(age.data)
choro.age$ggplot_scale <- scale_fill_brewer(name='Per cent old houses', palette = 'YlGn', drop=FALSE)
choro.age$set_zoom_zip(state_zoom =NULL, county_zoom=NULL, msa_zoom=NULL, zip_zoom = c('60607', '60608', '60609', '60610', '60611', '60612', '60613', '60614',
                            '60615', '60616', '60617', '60618', '60619', '60620', '60621', '60622',
                            '60623', '60624', '60625', '60626', '60628', '60629', '60630', '60827',
                            '60631', '60632', '60633', '60634', '60636', '60637', '60638', '60707',
                            '60639', '60640', '60641', '60643', '60644', '60645', '60646', '60647',
                            '60649', '60651', '60652', '60653', '60655', '60656', '60657'))

col1.4 <- insure$Zip
col2.4 <- insure$Invol
inv.data <- as.data.frame(cbind(col1.4,col2.4))
colnames(inv.data) <- c('region', 'value')
inv.data$region <- as.character(inv.data$region)
inv.data$value <- as.numeric(inv.data$value)
inv.data[18,1] <- '60827'
inv.data[45,1] <- '60707'

choro.inv <- ZipChoropleth$new(inv.data)
choro.inv$ggplot_scale <- scale_fill_brewer(name='FAIR plans per 100 houses', palette = 'YlOrBr', drop=FALSE)
choro.inv$set_zoom_zip(state_zoom =NULL, county_zoom=NULL, msa_zoom=NULL, zip_zoom = c('60607', '60608', '60609', '60610', '60611', '60612', '60613', '60614',
                            '60615', '60616', '60617', '60618', '60619', '60620', '60621', '60622',
                            '60623', '60624', '60625', '60626', '60628', '60629', '60630', '60827',
                            '60631', '60632', '60633', '60634', '60636', '60637', '60638', '60707',
                            '60639', '60640', '60641', '60643', '60644', '60645', '60646', '60647',
                            '60649', '60651', '60652', '60653', '60655', '60656', '60657'))

grid.arrange(choro.theft$render(), choro.fire$render(), choro.inc$render(), choro.race$render(), choro.age$render(), choro.inv$render(), ncol=2, nrow=3)
```

>Figure 3: Chicago Zip code maps for insurance data. Top left: Theft occurrences per 1000 housing units in Chicago, 1975. Top right: Fire incidents per 1000 housing units in Chicago, 1975. Centre left: Median family income census data. Centre right: Racial composition (in per cent minority). Bottom left: Percentage of housing units built in or before 1939. Bottom right: Data on involuntary market activity (number of FAIR plan policies and renewals per 100 housing units).

```{r, echo=FALSE,  message=FALSE, warning=FALSE}
ggplot(insure, aes(x = Income, y = Theft, colour = Race))+geom_point(shape = 19, size = 3, alpha = 0.6) +scale_colour_gradient(low = "orange", high = "brown")+labs(x = 'Median Household Income', y= 'Theft incidents Per 1000 Houses)')+ggtitle('Relationship between thefts, race and income')+theme_light()
```

>Figure 4: We can see from the above plot that theft incidences decrease as median household income increases. We also find that thefts occur more frequently in neighbourhoods where minority groups are more prevalent. 

```{r, echo=FALSE,  message=FALSE, warning=FALSE}
ggplot(insure, aes(x = Income, y = Volun, colour = Age))+geom_point(shape = 19, size = Race/10, alpha = 0.6) +scale_colour_gradient(low = "yellow3", high = "darkgreen")+labs(x = 'Median Household Income', y= 'Voluntary Market Activity)')+theme_light()
```

>Figure 5: Relationship active voluntary insure policies, race, age and income. We see that voluntary market activity shares a positive relationship with median household income. The area of the plotting circles corresponds to the per cent minority population, which highlights a general decrease in incomes and voluntary market activity as percentage increases. 
